Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2

### 2025-04-14 go3.tex
Penultimate before cleaning

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Explicit Scale Simulation for analysis of RNA-sequencing with
ALDEx2}
\author{true \and true \and true}
\date{}

@@

\doublespacing 
\singlespacing

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the actual size of the sampled DNA pool and the
scale (e.g., size, microbial load, or total gene expression) of the
measured biological system is degraded or lost. In the end, the
information contained in the data relates only to relative abundances
and has an arbitrary scale imposed by the sequencing process (1--3).
Researchers can use modified experimental protocols (4--7) or machine
learning methods (8) to uncover the biological variation in scale.
However, wet-lab protocols only provide information on the size of the
data downstream of the step in the sample preparation protocol where the
intervention was made and introduce an additional source of variation
that must be accounted for (5). The computational methods developed for
microbiome analysis have low correlation with the actual scale but are
useful (8). In short, the disconnect between sample-to-sample variation
in sequencing depth and biological variation in scale remains an open
challenge.

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR) (9--13),
exhibited as dataset-dependent FDR control and by the disconnect between
statistical and biological significance (14). In current practice, these
issues are addressed by a dual-filtering method, whereby both a low
p-value (or equivalently a low q-value following FDR correction (15))
and a large difference between groups is used to identify interesting
transcripts or genes for follow-up analysis (14, 16). This
double-filtering approach is graphically exemplified by the volcano plot
(16), but is known to not appropriately control the FDR (17, 18). The
second issue is poor performance when analyzing data where the mean
change between groups is non-zero (3). Such asymmetric data can arise
when a gene set is expressed in one group but not the other, or when one
group contains different gene content from the other. This type of data
frequently arises in in-vitro selection experiments (SELEX),
transcriptome analysis, and microbiome analysis (19). The third issue is
that the actual scale of the environment is often a major confounding
variable during analysis (3, 8). The final issue is that these problems
become more pronounced as more samples are collected; that is, more
information results in a worsening of the accuracy of the analysis (3,
13, 20).

The four problems were recently shown by Nixon et al. (3) to be a result
of a mismatch between the underlying size or scale of the system and the
assumptions of the normalizations used for the analysis of HTS.
Biological variation in scale often represents an important unmeasured
confounder in HTS analyses (21). For example, cells transformed by the
cMyc oncogene have about 3 times the amount of mRNA and about twice the
rRNA content than non-transformed cells (22), and this dramatically
skews transcriptome analysis unless spike-in approaches are used (5). In
addition, wild-type and mutant strains of cell lines, yeast or bacteria
have different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(23--25). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (26), and the
composition between these states is dramatically different (27, 28).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale (20).

Recently, Nixon et al. (3) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale but these implicit assumptions are
often inappropriate and difficult to interpret. This causes different
normalizations to provide different outputs when applied to the same
dataset (9, 11, 29--31). Intuitively, normalizations in widespread use
assume that either all samples have the same scale, e.g.~proportions,
rarefaction (32), RPKM (33, 34), etc; or that a subset of features in
one sample can be chosen as a reference to which the others are scaled
e.g.~the TMM (35), or LVHA (19) or the additive log-ratio (36); or that
different sub-parts of each sample maintain a constant scale across
samples e.g.~the RLE (37); or that the geometric mean of the parts is
appropriate e.g the CLR (38) and its derivatives.

The original naive ALDEx2 (39) model unwittingly made a strict
assumption about scale through the CLR normalization (3). This
assumption was often close enough to the true value to be useful, but
was not always the a good estimate and could be outperformed by other
normalizations (40). Nixon et al. (3) showed that better scale
assumptions resulted in more reproducible data analysis including better
control of both false positive and false negative results. We recently
modified ALDEx2 to explicitly model the scale over a range of reasonable
normalization parameters, and showed significant improvements in
performance in microbiome and in-vitro selection experiments (20). Here,
we briefly review these modifications and show how scale uncertainty can
greatly improve modeling in transcriptome and meta-transcriptome
datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given
in (3, 20). To be concrete, we let \(\mathbf{Y}\) denote the
\emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
of class \(d\) in the biological system from which sample \(n\) was
obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
or as the logarithm
\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
\mathbf{W}^{\perp}_{n}\).

Many of the normalizations in widespread use in tools such as DESeq2
(41), edgeR (35), metagenomeSeq (42) ALDEx2 (43) can be stated as ratios
of the form \(\hat{{\mathbf{W}}}_{dn}
\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is
determined by some function of the observation. We use the
\(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
\)) notation to indicate that the output is an estimate of the true
value. The technical variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (39, 43) is a general purpose toolbox to model the
uncertainty of HTS data and to use that model to estimate the underlying
LFC (log-fold change) significance. At a high-level, ALDEx2 has three
connected components to estimate the uncertainty inherent in HTS
datasets. First, the tool accounts for the uncertainty of the sequencing
counts using Dirichlet multinomial sampling to build a probabilistic
model of the data; i.e.,
\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
data (39). However, this step was modified recently to account for scale
uncertainty and misspecification (20) via a scale model, explained with
more details in (3, 20) and summarized in the next paragraph. Finally, a
standard null-hypothesis test and a non-parametric estimate of mean
standardized difference are used to report on the finite sample
variation. These sources of uncertainty and variation are combined via
reporting the expected values from a Monte-Carlo simulation framework.
For simplicity, we use the term `difference' to refer to the absolute
difference between groups, and `dispersion' to refer to the
within-condition difference or pooled variance as defined in (39). These
are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
20, 39, 43).

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
into a specialized type of statistical model which Nixon et al. (3) term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon
et al. (3) generalized the concept of normalizations by introducing the
concept of a \textit{scale model} to account for potential error in the
centred log-ratio normalization step. They did this by including a model
for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by
ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\),
where \(G_n\) is the geometric mean of sample n, which while being a
random variable, is essentially constant across each Monte-Carlo
replicate, but that differs between samples. With this modification,
ALDEx2 can be generalized by considering probability models for the
scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda
\sim N(\mu, \gamma^{2})\].

This formulation is quite flexible (3, 20). In the simple or `default'
configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
from a log-Normal distribution(3). Adding scale uncertainty with the
\(\gamma\) paramenter controls only the degree of uncertainty of the CLR
assumption for the \(x_{n}\) binary condition indicator (e.g.,
\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the
advanced or `informed' configuration, \(\mu\) takes different values for
each group and controls the location of the LFC assumption; combining
\(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the
location and the scale. An example of both the default and informed
approaches is given for a microbiome dataset in (20) showing increased
sensitivity and specificity. Here we show that these approaches also
work well in transcriptome and metatranscriptome datasets. These
modifications are instantiated in ALDEx2 which is the first software
package designed for SSRV-based inference.

@@

Gierliński et al. (44) generated a highly replicated yeast transcriptome
dataset to compare gene expression between a wild-type strain and a snf2
gene knockout, \(\Delta\)snf2. This dataset was used to test several
RNA-seq tools for their power to detect the set of differentially
abundant transcripts identified in the full dataset when the data was
subset (14). In this study each tool had its own `gold standard' set of
transcripts with different tools identifying between between 65\% to
\textgreater80\% of all transcripts as being significantly different.
Since the majority of transcripts were significantly different, the
authors suggested that it was more appropriate to apply a dual cutoff
composed of both a Benjamini-Hochberg (45) corrected p-value (q-value)
plus a difference cutoff to limit the number of identified transcripts
to a much smaller fraction of the total. Nixon et al, (20) showed that
adding even a small amount of scale uncertainty with ALDEx2 dramatically
reduced the number of significant transcripts identified, removing the
need for the dual cutoff approach in this dataset and others. Below we
include an intuitive explanation of why and how incorporating scale
uncertainty achieved this outcome using a setting of \(\gamma=0.5\). The
approach is in line with the recommendations of (20) and gives results
comparable to those proposed in (14).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}
\usepackage[left]{lineno}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Explicit Scale Simulation for analysis of RNA-sequencing count data with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Explicit Scale Simulation for analysis of RNA-sequencing count
data with ALDEx2}
\author{true \and true \and true}
\date{}

@@

\doublespacing \singlespacing 
\linenumbers

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the actual number of molecules in the sampled DNA
pool and the environmental scale (e.g., total number of molecules,
microbial load, or total gene expression) of the measured biological
system is degraded or lost. In the end, the information contained in the
data relates only to relative abundances and has an arbitrary scale
imposed by the sequencing process (1--3).

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale in the output
data. The first issue is poor control of the false discovery rate (FDR)
(4--8), exhibited as dataset-dependent FDR control which is observed as
a disconnect between statistical and biological significance (8, 9). In
current practice, this issue is addressed by a dual-filtering method,
whereby both a low p-value (or equivalently a low q-value following FDR
correction (10)) and a large difference between groups is used to
identify transcripts or genes of interest for follow-up analysis (9,
11). This double-filtering approach is graphically exemplified by the
volcano plot (11), but is known to not appropriately control the FDR
(12, 13). Since there is no standard way of determining what fold-change
cutoff should be used researchers have unlimited degrees of freedom
which is known to lead to unreliable inference (14). In particular Li et
al. (8) used patient or clinical-derived transcriptome datasets and
found that many methods suffer from an extremely high false positive
rate. The second issue is poor performance when analyzing data where the
mean change between groups is non-zero (3). Such asymmetric data can
arise when a gene set is expressed in one group but not the other, or
when one group contains different gene content from the other. This type
of data frequently arises in in-vitro selection experiments (SELEX),
transcriptome analysis, and microbiome analysis (15). The third issue is
that the actual scale of the environment is often a major confounding
variable during analysis (3, 16). This has long been known in
transcriptome analysis and was a major driver for the development of
normalization factors (17) and the use of molecular spike-ins to provide
a reference set of known counts (18--21) to estimate scale. While often
useful, spike-in methods only provide information downstream of the step
in the sample preparation protocol where the intervention was made and
introduce an additional source of variation that must be accounted for
(19). In the microbiome field, a recent landmark paper showed that
biological scale was a major unacknowledged confounder in many human
analyses (16). These authors built a machine learning model to uncover
the biological variation in scale and including this information was
useful despite exhibiting only a modest correlation (estimated to be
0.6) with the scale of data external to the training set (16). The final
issue is that all the above problems become more pronounced as more
samples are collected; that is, more information results in optimizing
for a precise but inaccurate analysis (3, 8, 22).

The four problems were recently shown by Nixon et al. (3) to be a result
of a mismatch between the underlying size or scale of the system and the
assumptions of the normalizations used for the analysis of HTS.
Biological variation in scale often represents an important unmeasured
confounder in HTS analyses (15, 16, 19, 23). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (24), and
this dramatically skews transcriptome analysis unless spike-in
approaches are used (19). In addition, wild-type and mutant strains of
cell lines, yeast or bacteria have different growth rates and RNA
contents under different conditions, which affect our ability to
identify truly differentially abundant genes (25--27). As another
example, the total bacterial load of the vaginal microbiome differs by
1-2 orders of magnitude in absolute abundance between the healthy and
bacterial vaginosis states (28), and the cell and RNA composition
between these states is dramatically different (29, 30). Thus, a full
description of any of these systems includes both relative change
(composition) and absolute abundance (scale). Current methods access
only the compositional information yet make implicit assumptions about
the scale (22).

Nixon et al. (3) showed that the challenge of non-biological variation
in sequencing depth could be explained as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale but these implicit assumptions are
often inappropriate and difficult to interpret. As a result, different
normalizations provide different outputs when applied to the same
dataset (4, 6, 17, 31, 32). Intuitively, normalizations in widespread
use assume that either all samples have the same scale,
e.g.~proportions, rarefaction (33), RPKM (34, 35), etc; or that a subset
of features in one sample can be chosen as a reference to which the
others are scaled e.g.~the TMM (36), the LVHA (15) qPCR (37), the
additive log-ratio (38); or that different sub-parts of each sample
maintain a constant scale across samples e.g.~the RLE (39); or that the
geometric mean of the parts is appropriate e.g the CLR (40) and its
derivatives. Notably when the assumption of similar scale across samples
is not violated, or violated only weakly, any method will provide a
reasonable analysis. The problem arises when this assumption is
violated, in which case all tools fail without warning.

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{./go3_files/figure-latex/scale-explained}

}

\caption{Mismatches between estimated scale and true scale lead to poor estimation of high throughput sequencing data. All normalizations used for differential abundance analysis make some strict assumption about the scale of the environment as shown in line 1. In this example all normalizations produce a biased estimate of the environmental scale, but estimate C is the closest to the truth.  Adding uncertainty to normalization C as represented by the distribution in the line 2 leads to less bias and now includes the actual environmental scale in the assumption. As shown in line 3 in some cases it may be useful to adjust the centre of the scale uncertainty estimate if the initial normalizations give very poor estimates of the underlying environment.}\label{fig:unnamed-chunk-1}
\end{figure}

The original scale-naive ALDEx2 (41) model unwittingly made a strict
assumption about scale through the CLR normalization and we found that
the CLR was very sensitive to violations of the assumption of scale
identity between groups (15). Moreover, when the assumption of identity
was not true the CLR used by ALDEx2 could be outperformed by other
normalizations in simulation studies (42). As illustrated graphically in
Figure 1, Nixon et al. (3) showed through simulation that introducing
uncertainty in the scale assumptions, and in extreme cases altering the
location of the scale assumption, resulted in more reproducible data
analysis including better control of both false positive and false
negative results. We modified ALDEx2 to explicitly model uncertainty in
scale over a range of reasonable normalization parameters, and showed
significant improvements in performance in microbiome and in-vitro
selection experiments (22) and in a vaginal metatranscriptome analysis
(43). Here, we briefly review these modifications and show how adding
scale uncertainty can greatly improve modeling in transcriptome and
meta-transcriptome datasets to provide substantially more robust and
reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given
in (3, 22). To be concrete, we let \(\mathbf{Y}\) denote the
\emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
\(n\). Likewise, we can denote \(\mathbf{W}_{dn}\) as the \emph{true}
amount of class \(d\) in the biological system from which sample \(n\)
was obtained. We can think of \(\mathbf{W}\) as consisting of two parts,
the scale \(\mathbf{W}^{Tot}\) (e.g., totals) and the composition
\(\mathbf{W}^{Comp}\) (i.e., proportions). That is, \(\mathbf{W}^{Tot}\)
is a \(N\)-vector with elements
\(\mathbf{W}^{Tot}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{Comp}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{Comp}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{Tot}_{n}\). Note
that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{Comp}_{dn}\mathbf{W}^{Tot}_{n}\), or as
the logarithm
\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{Comp}_{dn} + \log \mathbf{W}^{Tot}_{n}\).

Many of the normalizations used in tools such as DESeq2 (44), edgeR
(36), metagenomeSeq (45) ALDEx2 (46) can be stated as ratios of the form
\(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
the denominator is determined by some function of the observation. We
use the hat notation (\(\ \hat{}\ \)) to indicate that the output is an
estimate of the true value. The technical variation in sequencing depth,
which is often called ``library size'' has no relationship with the
actual number of molecules in the sampled environment (1). In other
words (\(\mathbf{Y}^{Tot}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) the observed
data \(\mathbf{Y}\) provides us with information about the system
composition \(\mathbf{W}^{Comp}\) but little to no information in the
system scale \(\mathbf{W}^{Tot}\) (Lovell et al.~2011).

@@

The ALDEx2 R package (41, 46) is a general purpose toolbox to model the
uncertainty of HTS data and to use that model to estimate the
significance of the underlying LFC (log-fold change). At a high-level,
ALDEx2 has three connected components to estimate the uncertainty
inherent in HTS datasets. First, the tool accounts for the uncertainty
of the sequencing counts using Dirichlet multinomial sampling to build a
probabilistic model of the data; i.e.,
\(\mathbf{\hat{W}}^{Comp} \approx \mathrm{Dir}(\mathbf{Y})\). Secondly,
ALDEx2 uses the centred log-ratio transformation to scale the data (41).
It was this step that was modified to account for scale uncertainty and
misspecification (22) explained with more details in (3, 22) and
summarized in the next paragraph. Finally, a standard null-hypothesis
test and a non-parametric estimate of mean standardized difference are
used to report on the finite sample variation. These sources of
uncertainty and variation are combined via reporting the expected values
from a Monte-Carlo simulation framework. For simplicity, we use the term
`difference' to refer to the absolute difference between groups, and
`dispersion' to refer to the within-condition difference or pooled
variance as defined in (41). These are calculated on a \(\log_2\) scale.
For more details on ALDEx2 see (3, 22, 41, 46).

Scale models were incorporated into ALDEx2, turning the ALDEx2 model
into a specialized type of statistical model which Nixon et al. (3) term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon
et al. (3) generalized the concept of normalizations by introducing the
concept of a \textit{scale model} to account for potential error in the
centred log-ratio normalization step. They did this by including a model
for \(\mathbf{\hat{W}}^{Tot}_{n}\). The CLR normalization used by ALDEx2
makes the assumption \(\mathbf{\hat{W}}^{Tot}_{n}=1/G_{n}\), where
\(G_n\) is the geometric mean of the counts (or the corresponding
proportions)of each part in sample n, which while being a random
variable, is essentially constant across each Monte-Carlo replicate, but
that differs between samples. With this modification, ALDEx2 can be
generalized by considering probability models for the scale
\(\mathbf{\hat{W}}^{Tot}_{n}\) that have mean \(1/G_{n}\). For example,
the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{Tot}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(\mu, \gamma^{2}).\]

This formulation is quite flexible (3, 22). In the simple or `default'
configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
from a log-Normal distribution (3). Adding scale uncertainty with the
\(\gamma\) paramenter (as shown in Figure 1:1) controls only the degree
of uncertainty of the CLR assumption for the \(x_{n}\) binary condition
indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes
control). In the advanced or `informed' configuration, \(\mu\) takes
different values for each group and controls the location of the LFC
assumption (as shown in Figure 1:2); combining \(\mu\) with a \(\gamma\)
estimate allows for uncertainty in both the location and the scale. In
the manuscript where the idea of scale was original derived, Nixon et
al. (3) conducted extensive simulations showing that both the default
and informed approaches exhibit increased sensitivity and specificity
(22). In this report we show that these approaches also work well to
control the FDR in transcriptome and metatranscriptome datasets. We also
provide some additional insights into how this is achieved. These
modifications are instantiated in ALDEx2 which is the first software
package designed for SSRV-based inference.

@@

Gierliński et al. (47) generated a highly replicated yeast transcriptome
dataset to compare gene expression between a wild-type strain and a snf2
gene knockout, \(\Delta\)snf2. This dataset of 86 samples (44 and 42 per
group) is an example of technical growth replicate experiments common in
the literature. The dataset was used to test several RNA-seq tools for
their power to detect the set of differentially abundant transcripts
identified in the full dataset when the data was subset (9). In this
original study each tool had its own `gold standard' set of transcripts
with different tools identifying between between 65\% to
\textgreater80\% of all transcripts as being significantly different.
Since the majority of transcripts were significantly different, the
authors suggested that it was more appropriate to apply a dual cutoff
composed of both a Benjamini-Hochberg (48) corrected p-value (q-value)
plus a difference cutoff to limit the number of identified transcripts
to a much smaller fraction of the total. In other datasets, Nixon et al,
(22) showed that adding even a small amount of scale uncertainty with
ALDEx2 dramatically reduced the number of significant transcripts
identified, removing the need for the dual cutoff approach in this
dataset and others.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{go3_files/figure-latex/yst-plot-1}

}

\caption{Volcano and effect plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential expression (DE) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots or difference and the q-values using volcano plots. Panels A,C are for the naive analyses, and Panels B,D are for the  default analyses that include scale uncertainty. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the both analyses (red) or in the naive analysis only (orange). Points in grey are not statistically signficantly different under any condition. The horizontal dashed lines represent a \(\log_2\)difference of \(\pm 1.4\).}\label{fig:yst-plot}
\end{figure}
